---
title: "Gwas"
output:
  html_document:
    df_print: paged
---

# Load in the GWASpoly library

```{r}
library(GWASpoly)
```


### Load PCA and DAPC structure matrices

```{r}
dir.create("GWAS", showWarnings = FALSE)
dir.create("GWAS/Yield", showWarnings = FALSE, recursive = TRUE)
pca <- read.csv("DAPC_PIC_output/PCA_principal_components.csv", header=TRUE)
dapc <- read.csv("DAPC_PIC_output/DAPC_membership_probabilities.csv", header=TRUE)
```


### Yield with PCA
First, the phenotype data needs to be matched to the PCA data.

```{r}
geno <- read.csv("sharma_genotype_GWAS_poly.csv", header=TRUE)
pheno <- read.csv("BLUEs_PIC_output/BLUES.csv", header=TRUE)
pheno_yield <- pheno[, c("ID", "yield_kg")]
colnames(pheno_yield) <- c("Sample", "Yield")
write.csv(pheno_yield, "GWAS/Yield/pheno_yield.csv", row.names=FALSE)
```
```{r}
colnames(pca)[1] <- "Sample"
```

```{r}

merged_yield <- merge(pca, pheno_yield, by = "Sample")
merged_yield <- merged_yield[, c("Sample", "Yield", setdiff(names(merged_yield), c("Sample", "Yield")))]
write.csv(merged_yield, file = "GWAS/Yield/merged_yield.csv", row.names = FALSE)

#Check merged_yield file matches the phenotype and PCA data to the correct sample IDs


```

Then, GWASpoly for Yield with PCA data can be performed.

```{r}
#Load data into GWASpoly
data_yield_PCA <- read.GWASpoly(
  ploidy = 4,
  pheno.file = "GWAS/Yield/merged_yield.csv",
  geno.file = "sharma_genotype_GWAS_poly.csv",
  format = "numeric",  
  n.traits=1,
  delim = ","
)


#Set population covariates for PCA
params_yield_PCA <- set.params(
  fixed = c("PC1", "PC2", "PC3", "PC4", "PC5"),
  fixed.type = rep("numeric", 5),
  n.PC = 5,
  MAF = NULL,
  geno.freq = NULL,
  P3D = TRUE
)


#Set K matrix
data_yield_PCA <- set.K(data_yield_PCA, K = NULL, n.core = 1, LOCO = FALSE)


#Run GWASpoly for yield PCA
gwas_yield_pca <- GWASpoly(
  data = data_yield_PCA,
  models = c("additive", "general", "1-dom", "2-dom"),
  traits = "Yield",
  params = params_yield_PCA,
  n.core = 4  
)


#Set significant threshold
gwas_yield_pca <- set.threshold(
  data = gwas_yield_pca,
  method = "Bonferroni",
  level = 0.05,
  n.core = 4
)

```

Once GWASpoly has been completed, Manhattan plots and QQ plots can be made. 

```{r}
#Manhattan plots
pdf("GWAS/Yield/manhattan_yield_PCA_additive.pdf", width = 8, height = 6)
manhattan.plot(gwas_yield_pca, traits="Yield", models="additive")
dev.off()

pdf("GWAS/Yield/manhattan_yield_PCA_general.pdf", width = 8, height = 6)
manhattan.plot(gwas_yield_pca, traits="Yield", models="general")
dev.off()

pdf("GWAS/Yield/manhattan_yield_PCA_1domalt.pdf", width = 8, height = 6)
manhattan.plot(gwas_yield_pca, traits="Yield", models="1-dom-alt")
dev.off()

pdf("GWAS/Yield/manhattan_yield_PCA_1domref.pdf", width = 8, height = 6)
manhattan.plot(gwas_yield_pca, traits="Yield", models="1-dom-ref")
dev.off()

pdf("GWAS/Yield/manhattan_yield_PCA_2domalt.pdf", width = 8, height = 6)
manhattan.plot(gwas_yield_pca, traits="Yield", models="2-dom-alt")
dev.off()

pdf("GWAS/Yield/manhattan_yield_PCA_2domref.pdf", width = 8, height = 6)
manhattan.plot(gwas_yield_pca, traits="Yield", models="2-dom-ref")
dev.off()


#QQ plot
pdf("GWAS/Yield/QQ_yield_PCA.pdf", width = 8, height = 6)
qq.plot(gwas_yield_pca, trait = "Yield")
dev.off()
```

The next step is to create an SNP size effect table using the GWASpoly results.

```{r}
models <- c("additive", "general", "1-dom-alt", "1-dom-ref", "2-dom-alt", "2-dom-ref")

gwas_yield_PCA_results <- data.frame()

#Loop over models
for (model in models) {
  #Get the threshold for the current model
  threshold <- gwas_yield_pca@threshold["Yield", model]
  
  #Create a temporary results table for the current model
  temp_results <- data.frame(
    SNP = gwas_yield_pca@map$Marker,
    Chromosome = gwas_yield_pca@map$Chrom,
    Position = gwas_yield_pca@map$Position,
    Model = model,
    Effect = gwas_yield_pca@effects$Yield[[model]],
    Pvalue = gwas_yield_pca@scores$Yield[[model]],
    Significant = gwas_yield_pca@scores$Yield[[model]] >= threshold
  )
  
  #Add to final results
  gwas_yield_PCA_results <- rbind(gwas_yield_PCA_results, temp_results)
}

# Save as CSV
write.csv(
  gwas_yield_PCA_results,
  "GWAS/Yield/yield_PCA_snp_effect_table.csv",
  row.names = FALSE
)
```

Using this file, filter the Significant column for 'TRUE' values only, and use these for compiling a table for the significant SNPs for Yield. GWASpoly for Yield using PCA found 6 significant SNPs across chromosomes 6 and 7.



### Yield with DAPC
The phenotype data for Yield needs to be matched to the DAPC data. The initial pheno_yield.csv file has already been created in ### Yield with PCA.

```{r}
# Rename the first column of the DAPC data frame to 'Sample'
colnames(dapc)[1] <- "Sample"

merged_yield_DAPC <- merge(dapc, pheno_yield, by = "Sample")
merged_yield_DAPC <- merged_yield_DAPC[, c("Sample", "Yield", setdiff(names(merged_yield_DAPC), c("Sample", "Yield")))]
write.csv(merged_yield_DAPC, file = "GWAS/Yield/merged_yield_DAPC.csv", row.names = FALSE)

#Check merged_yield_DAPC file matches the phenotype and DAPC data to the correct sample IDs

```

Then, GWASpoly for Yield with DAPC data can be performed.

```{r}
#Load data into GWASpoly
data_yield_DAPC <- read.GWASpoly(
  ploidy = 4,
  pheno.file = "GWAS/Yield/merged_yield_DAPC.csv",
  geno.file = "sharma_genotype_GWAS_poly.csv", # The corrected path
  format = "numeric",  
  n.traits=1,
  delim = ","
)


#Set population covariates for DAPC
params_yield_DAPC <- set.params(
  fixed = c("X1", "X2", "X3", "X4"),
  fixed.type = rep("numeric", 4),
  n.PC = 4,
  MAF = NULL,
  geno.freq = NULL,
  P3D = TRUE
)


#Set K matrix
data_yield_DAPC <- set.K(data_yield_DAPC, K = NULL, n.core = 1, LOCO = FALSE)


#Run GWAS for yield DAPC
gwas_yield_DAPC <- GWASpoly(
  data = data_yield_DAPC,
  models = c("additive", "general", "1-dom", "2-dom"),
  traits = "Yield",
  params = params_yield_DAPC,
  n.core = 4 
)


#Set significant threshold
gwas_yield_DAPC <- set.threshold(
  data = gwas_yield_DAPC,
  method = "Bonferroni",
  level = 0.05,
  n.core = 4
)
```

Once GWASpoly has been completed, Manhattan plots and QQ plots can be made. 

```{r}
#Manhattan plots
pdf("GWAS/Yield/manhattan_yield_DAPC_additive.pdf", width = 8, height = 6)
manhattan.plot(gwas_yield_DAPC, traits="Yield", models="additive")
dev.off()

pdf("GWAS/Yield/manhattan_yield_DAPC_general.pdf", width = 8, height = 6)
manhattan.plot(gwas_yield_DAPC, traits="Yield", models="general")
dev.off()

pdf("GWAS/Yield/manhattan_yield_DAPC_1domalt.pdf", width = 8, height = 6)
manhattan.plot(gwas_yield_DAPC, traits="Yield", models="1-dom-alt")
dev.off()

pdf("GWAS/Yield/manhattan_yield_DAPC_1domref.pdf", width = 8, height = 6)
manhattan.plot(gwas_yield_DAPC, traits="Yield", models="1-dom-ref")
dev.off()

pdf("GWAS/Yield/manhattan_yield_DAPC_2domalt.pdf", width = 8, height = 6)
manhattan.plot(gwas_yield_DAPC, traits="Yield", models="2-dom-alt")
dev.off()

pdf("GWAS/Yield/manhattan_yield_DAPC_2domref.pdf", width = 8, height = 6)
manhattan.plot(gwas_yield_DAPC, traits="Yield", models="2-dom-ref")
dev.off()


#QQ plot
pdf("GWAS/Yield/QQ_yield_DAPC.pdf", width = 8, height = 6)
qq.plot(gwas_yield_DAPC, trait = "Yield")
dev.off()
```

Then, create an SNP size effect table using the GWASpoly results.

```{r}
models <- c("additive", "general", "1-dom-alt", "1-dom-ref", "2-dom-alt", "2-dom-ref")

gwas_yield_DAPC_results <- data.frame()

#Loop over models
for (model in models) {
  #Get the threshold for the current model
  threshold <- gwas_yield_DAPC@threshold["Yield", model]
  
  #Create a temporary results table for the current model
  temp_results <- data.frame(
    SNP = gwas_yield_DAPC@map$Marker,
    Chromosome = gwas_yield_DAPC@map$Chrom,
    Position = gwas_yield_DAPC@map$Position,
    Model = model,
    Effect = gwas_yield_DAPC@effects$Yield[[model]],
    Pvalue = gwas_yield_DAPC@scores$Yield[[model]],
    Significant = gwas_yield_DAPC@scores$Yield[[model]] >= threshold
  )
  
  #Add to final results
  gwas_yield_DAPC_results <- rbind(gwas_yield_DAPC_results, temp_results)
}

# Save as CSV
write.csv(
  gwas_yield_DAPC_results,
  "GWAS/Yield/yield_DAPC_snp_effect_table.csv",
  row.names = FALSE
)
```

Using this file, filter the Significant column for 'TRUE' values only, and use these for compiling a table for the significant SNPs for Yield. GWASpoly for Yield using DAPC also found 6 significant SNPs across chromosomes 6 and 7.

As the Manhattan plot for the General DAPC model shows the most SNPs above the threshold value, this model was chosen for creating boxplots for the significant SNPs in Yield.

To create these boxplots, ggplot2 is used. 

To extract the -log10 values, the genotype data first needs to be transposed so that rows show the samples and columns show the SNPs.

```{r}
#Transpose genotype data
geno_raw <- read.csv("sharma_genotype_GWAS_poly.csv", row.names = 1, check.names = FALSE)
geno_t <- t(geno_raw)  # transpose: now rows=samples, columns=SNPs
geno_t <- as.data.frame(geno_t)


#Call DAPC data for the phenotype
pheno <- read.csv("GWAS/Yield/merged_yield_DAPC.csv", stringsAsFactors = FALSE)


#Match samples present in both genotype and phenotype
common_samples <- intersect(rownames(geno_t), pheno$Sample)
geno_t <- geno_t[common_samples, , drop = FALSE]
pheno <- pheno[pheno$Sample %in% common_samples, ]
pheno <- pheno[match(common_samples, pheno$Sample), ]


#Specify trait and model
trait <- "Yield"
model <- "general"


#Get Bonferroni threshold for the specified trait and model
bonf_threshold <- gwas_yield_DAPC@threshold[trait, model]


#Extract -log10(p-values) for the trait/model
logpvals <- gwas_yield_DAPC@scores[[trait]][[model]]


#Significant SNPs are those with -log10(pval) > threshold
sig_indices <- which(logpvals > bonf_threshold)


#Extract significant SNPs
y_sig_qtls <- data.frame(
  Marker = gwas_yield_DAPC@map$Marker[sig_indices],
  Chrom = gwas_yield_DAPC@map$Chrom[sig_indices],
  Position = gwas_yield_DAPC@map$Position[sig_indices],
  log10_P = logpvals[sig_indices]
)


#Get SNP names (Markers)
y_sig_snps <- trimws(y_sig_qtls$Marker)


#Check which significant SNPs are in the genotype data
sig_snps_present <- y_sig_snps[y_sig_snps %in% colnames(geno_t)]


if (length(sig_snps_present) == 0) {
  stop("No significant SNPs found in genotype data columns.")
} else {
  message(length(sig_snps_present), " significant SNP(s) found in genotype data.")
}


#Load ggplot2 library
library(ggplot2)


#Create boxplot
plot_qtl_boxplot <- function(snp_id, geno_mat, pheno_df, trait) {
  geno_vec <- geno_mat[[snp_id]]
  pheno_vec <- pheno_df[[trait]]
  
  df <- data.frame(Genotype = as.factor(geno_vec), Phenotype = pheno_vec)
  
   genotype_labels <- c("0" = "AAAA", "1" = "AAAB", "2" = "AABB", "3" = "ABBB", "4" = "BBBB")
  
  ggplot(df, aes(x = Genotype, y = Phenotype)) +
    geom_boxplot() +
    labs(title = paste("QTL boxplot for Yield"),
         x = paste( "Genotype for", snp_id), 
         y = trait) +
    scale_x_discrete(labels = genotype_labels) +
    theme_minimal()
}

#Save boxplot as PDF
pdf("GWAS/Yield/QTL_boxplot_yield.pdf", width = 7, height = 5)
for (snp in sig_snps_present) {
  print(plot_qtl_boxplot(snp, geno_t, pheno, "Yield"))
}
dev.off()
```

Using the DAPC general model again, an improved QQ plot showing the significant SNPs only can be created. Significant SNPs were found on chromosomes 6 and 7, so two QQ plots are needed - one for each chromosome.

```{r}
# Extract scores for general model and map info, combine with SNP info
y_general_scores <- gwas_yield_DAPC@scores[["Yield"]][["general"]]
y_map_info <- as.data.frame(gwas_yield_DAPC@map)
y_results_df <- cbind(y_map_info, Score = y_general_scores)

# Ensure Chrom is character and clean common prefixes
y_results_df$Chrom <- as.character(y_results_df$Chrom)
y_results_df$Chrom <- gsub("^chr", "", y_results_df$Chrom, ignore.case = TRUE)
y_results_df$Chrom <- gsub("^Chr", "", y_results_df$Chrom, ignore.case = TRUE)
y_results_df$Chrom <- gsub("^0+", "", y_results_df$Chrom)  # remove leading zeros

# Function to make QQ plot for a chromosome
plot_chr_qq <- function(df, chr_num, out_file) {
  chr_df <- subset(df, Chrom == as.character(chr_num))
  
  if (nrow(chr_df) == 0) {
    warning(paste("No markers found for chromosome", chr_num))
    return(NULL)
  }
  
  # Convert -log10(p) scores back to p-values
  pvals <- 10^(-chr_df$Score)
  
  # Expected vs observed
  expected <- -log10(ppoints(length(pvals)))
  observed <- -log10(sort(pvals))
  
  # Save plot
  pdf(out_file, width = 7, height = 7)
  plot(expected, observed,
       main = paste("QQ Plot: Yield General Model - Chromosome", chr_num),
       xlab = "Expected -log10(p)",
       ylab = "Observed -log10(p)",
       pch = 20, col = "red")
  abline(0, 1, col = "black")
  dev.off()
}

# Run for chromosomes 6 and 7
plot_chr_qq(y_results_df, 6, "GWAS/Yield/QQplot_Yield_general_chr6.pdf")
plot_chr_qq(y_results_df, 7, "GWAS/Yield/QQplot_Yield_general_chr7.pdf")

```

These QQ plots will be used in the Results figures. 



### Yield with STRUCTURE
The phenotype data for Yield needs to be matched to the STRUCTURE data. The initial pheno_yield.csv file has already been created in ### Yield with PCA.

First, the k3 outfile needs to be converted into a csv file.

```{r}
lines <- readLines("K3.outfile")
lines <- lines[lines != ""]

#Process each line and convert to a dataframe
structure_data_list <- lapply(lines, function(line) {
    parts <- strsplit(line, ":")[[1]]
  left_fields <- strsplit(trimws(parts[1]), "\\s+")[[1]]
    right_fields <- strsplit(trimws(parts[2]), "\\s+")[[1]]

  data.frame(
    SampleID = left_fields[2],  # Sample ID
    Q1 = as.numeric(right_fields[1]),
    Q2 = as.numeric(right_fields[2]),
    Q3 = as.numeric(right_fields[3])
  )
})

#Combine all rows
structure_df <- do.call(rbind, structure_data_list)

#Save as csv
write.csv(structure_df, "GWAS/structure.csv", row.names = FALSE)

#Check structure.csv file matches the data in the k3 outfile

```

Next, the STRUCTURE data can be merged with the phenotype data.

```{r}
pheno_yield <- read.csv("GWAS/Yield/pheno_yield.csv")
structure <- read.csv("GWAS/structure.csv")

#Add number ID to phenotype
pheno_yield$STRUCTURE_ID <- 1:nrow(pheno_yield)

#Merge on ID
merged <- merge(pheno_yield, structure, by.x = "STRUCTURE_ID", by.y = "SampleID")
merged$STRUCTURE_ID <- NULL

#Save merged file
write.csv(merged, "GWAS/Yield/merged_yield_STRUCTURE.csv", row.names = FALSE)

#Check merged_yield_STRUCTURE file matches the phenotype and STRUCTURE data to the correct sample IDs
```

Then, GWASpoly for Yield with STRUCTURE data can be performed.

```{r}
#Load data into GWASpoly
data_yield_STRUCTURE <- read.GWASpoly(
  ploidy = 4,
  pheno.file = "GWAS/Yield/merged_yield_STRUCTURE.csv",
  geno.file = "sharma_genotype_GWAS_poly.csv",
  format = "numeric",  
  n.traits=1,
  delim = ","
)


#Set population covariates for STRUCTURE
params_yield_STRUCTURE <- set.params(
  fixed = c("Q1", "Q2", "Q3"),
  fixed.type = rep("numeric", 3),
  n.PC = 3,
  MAF = NULL,
  geno.freq = NULL,
  P3D = TRUE
)


#Set K matrix
data_yield_STRUCTURE <- set.K(data_yield_STRUCTURE, K = NULL, n.core = 1, LOCO = FALSE)


#Run GWAS for yield STRUCTURE
gwas_yield_STRUCTURE <- GWASpoly(
  data = data_yield_STRUCTURE,
  models = c("additive", "general", "1-dom", "2-dom"),
  traits = "Yield",
  params = params_yield_STRUCTURE,
  n.core = 4 
)


#Set significant threshold
gwas_yield_STRUCTURE <- set.threshold(
  data = gwas_yield_STRUCTURE,
  method = "Bonferroni",
  level = 0.05,
  n.core = 4
)
```

Once GWASpoly has been completed, Manhattan and QQ plots can be made.

```{r}
#Manhattan plots
pdf("GWAS/Yield/manhattan_yield_STRUCTURE_additive.pdf", width = 8, height = 6)
manhattan.plot(gwas_yield_STRUCTURE, traits="Yield", models="additive")
dev.off()

pdf("GWAS/Yield/manhattan_yield_STRUCTURE_general.pdf", width = 8, height = 6)
manhattan.plot(gwas_yield_STRUCTURE, traits="Yield", models="general")
dev.off()

pdf("GWAS/Yield/manhattan_yield_STRUCTURE_1domalt.pdf", width = 8, height = 6)
manhattan.plot(gwas_yield_STRUCTURE, traits="Yield", models="1-dom-alt")
dev.off()

pdf("GWAS/Yield/manhattan_yield_STRUCTURE_1domref.pdf", width = 8, height = 6)
manhattan.plot(gwas_yield_STRUCTURE, traits="Yield", models="1-dom-ref")
dev.off()

pdf("GWAS/Yield/manhattan_yield_STRUCTURE_2domalt.pdf", width = 8, height = 6)
manhattan.plot(gwas_yield_STRUCTURE, traits="Yield", models="2-dom-alt")
dev.off()

pdf("GWAS/Yield/manhattan_yield_STRUCTURE_2domref.pdf", width = 8, height = 6)
manhattan.plot(gwas_yield_STRUCTURE, traits="Yield", models="2-dom-ref")
dev.off()


#QQ plot
pdf("GWAS/Yield/QQ_yield_STRUCTURE.pdf", width = 8, height = 6)
qq.plot(gwas_yield_STRUCTURE, trait = "Yield")
dev.off()
```

Then, create an SNP size effect table using the GWASpoly results.

```{r}
models <- c("additive", "general", "1-dom-alt", "1-dom-ref", "2-dom-alt", "2-dom-ref")

gwas_yield_STRUCTURE_results <- data.frame()

#Loop over models
for (model in models) {
  #Get the threshold for the current model
  threshold <- gwas_yield_STRUCTURE@threshold["Yield", model]
  
  #Create a temporary results table for the current model
  temp_results <- data.frame(
    SNP = gwas_yield_STRUCTURE@map$Marker,
    Chromosome = gwas_yield_STRUCTURE@map$Chrom,
    Position = gwas_yield_STRUCTURE@map$Position,
    Model = model,
    Effect = gwas_yield_STRUCTURE@effects$Yield[[model]],
    Pvalue = gwas_yield_STRUCTURE@scores$Yield[[model]],
    Significant = gwas_yield_STRUCTURE@scores$Yield[[model]] >= threshold
  )
  
  #Add to final results
  gwas_yield_STRUCTURE_results <- rbind(gwas_yield_STRUCTURE_results, temp_results)
}

# Save as CSV
write.csv(
  gwas_yield_STRUCTURE_results,
  "GWAS/Yield/yield_STRUCTURE_snp_effect_table.csv",
  row.names = FALSE
)
```

Using this file, filter the Significant column for 'TRUE' values only, and use these for compiling a table for the significant SNPs for Yield. GWASpoly for Yield using STRUCTURE also found 6 significant SNPs across chromosomes 6 and 7.

```{r}
# Load the SNP effect table
gwas_yield_STRUCTURE_results <- read.csv("GWAS/Yield/yield_STRUCTURE_snp_effect_table.csv", stringsAsFactors = FALSE)

# Filter for significant SNPs (Significant == TRUE)
significant_snps <- gwas_yield_STRUCTURE_results[gwas_yield_STRUCTURE_results$Significant == TRUE, ]

# Select relevant columns for the table
significant_snps_table <- significant_snps[, c("SNP", "Chromosome", "Position", "Model", "Effect", "Pvalue", "Significant")]

# Sort by Chromosome and Position for clarity
significant_snps_table <- significant_snps_table[order(significant_snps_table$Chromosome, significant_snps_table$Position), ]

# Display the table
print(significant_snps_table)

# Optionally, save the table to a CSV file
write.csv(significant_snps_table, "GWAS/Yield/significant_snps_yield_STRUCTURE.csv", row.names = FALSE)
```

